The five pillars of computational reproducibility: Bioinformatics and beyond

### 2023-08-21 manuscript/rebuttal.Rmd

RESPONSE: The passage in question has been replaced, with reference to evidence from a large
survey that found that ~69% of university academics used MS Excel in their research activities.

> Spreadsheet errors could be widespread, given that it is used as an analysis tool by ~69% of
> researchers according to a survey undertaken in 2015-16 of 20,000 university academics @Kramer2016-fj.

> 6. Page 3, Line 56: It is not necessarily true that analyses performed using web tools are not
> reproducible. Although rare, some web tools facilitate reproducibility by providing code or
> configuration files and/or allowing the apps to be executed locally.

RESPONSE: Thank you for raising this point. We have amended the section in question:

> While webtools are valuable for data exploration, there are worries that they undermine
> reproducibility for the sake of convenience [@Wijesooriya2022-ly].
> Transferring data between compute platforms is also discouraged.
> For example, having workflows that involve combinations of web-based and scripted tools require
> data transfer steps which are inefficient and error-prone.
> On the other hand, some web based tools excel at reproducibility.
> The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
> bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
> [@The_Galaxy_Community2022-dt;@Reich2006-na].
> Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
> or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).

@@

> 7. Page 5, Line 15, type-o = "authors provided along the"

RESPONSE: The passage in question has been amended:

> In this work, authors provided alongside the traditional research paper a Jupyter notebook ..

@@

> 8. Page 6, Line 21: "quantum leap" (this term is overly optimistic in this context)

RESPONSE: The passage in question has been amended:

> These features make literate programming a useful tool for science communication in a range of
> situations.

> 9. How do you get from a notebook to an actual paper submission if you have to do custom formatting
> of the document, including references?
> My understanding is that this is still not possible, but please correct me if I'm wrong.

RESPONSE: It is possible to achieve accurate journal-specific formatting of references with a Rmarkdown
document.
It requires a few things:

* A .bib file that contains the bibliography in bibtex format.

* A .csl file that specifies the citation style.
These are avaialble from [GitHub](https://github.com/citation-style-language/styles).

* Modifcation to the Rmd header section that specifies the path of the bib and csl files.

This enables convenient reformatting of the bibliography for different venues.
In terms of overall document formatting, the `rticles` package provides a suite of
custom R Markdown LaTeX formats suitable for many journals.

We have amended the Literate Programming section to mention these features briefly:

> Quarto and R Markdown have the ability to automatically generate a bibliography in many different
> journal styles [@noauthor_undated-pg].
> Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
> freely available @Xie2018-rf.

@@

> 10. What about when your data files are too large to fit on a personal computer?

RESPONSE: When data files are too large for a PC, it is common to use high performance computers, shared
clusters or cloud computers.
These computers are typically accessed via SSH (headless mode), so Quarto and Rstudio graphical
interfaces might not be available.
In those cases, the commands `quarto render document1.qmd` and `rmarkdown::render("document2.Rmd")` will
work for Quarto and R Markdown respectively.

> 11. What about computationally intensive tasks that must be performed using specialized computing
> environments like the cloud or clusters?

RESPONSE: We have amended the relevant section to suggest a specific workflow manager for compute
intensive applications.

> Projects involving computationally intensive tasks might benefit from a build/workflow automation
> solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
> `nextflow` [@Di_Tommaso2017-yo].

@@

> 12. Page 6, line 32: The master script idea was already mentioned earlier.

RESPONSE: True. We have amended the "Documentation" section to remove the unnecessary reiteration:

> As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
> minimising the number of commands required.

> 13. Section on version control: This section focuses mostly on using VC for software development
> (that is my interpretation).
> To be consistent with the introduction, it should focus more on data analyses.
> Although I use VC for analyses, I feel that simpler approaches are better in many cases.
> For example Dropbox and Google Drive provide some version-control and backup functionality and do
> not require the same level of knowledge as git.

### 2023-08-21 manuscript/5pillars.Rmd
progress on reviewer 3

While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.

To facilitate an end-to-end automated process, the code and data need to be "linked", which means the
code is aware of the location of the data and how to get it automatically @Peng2011-ia.
The code should be able to fetch the data from a publicly accessible location and conduct the downstream
processing and analysis without having to acquire the data some other way.

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
Using free and open source software ensures that the research is accessible to the greatest audience
@Garijo2013-zl, as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects entailing big data analysis might benefit from a build/workflow automation solution like `make`,
`snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
need to be repeated.
After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
which saves labor and compute time.

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided along the traditional research paper a
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
that performed and documented the entire analysis procedure.
The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
Outputs include HTML documents, but also slideshows and even ebooks.

These features make literate programming a quantum leap for science communication in a range of
situations.
Whether this is sharing a data analysis report, giving a presentation at a meeting,
writing a research article or self-publishing an ebook, literate programming provides the
ability to construct transparent data narratives with clear provenance in a conveniently shareable
form.

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, and using a master
script is a simple way to do this.
It also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux installation [@Ziemann2023-nm].
By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.

>>>>

While webtools are valuable for data exploration, there are worries that they undermine
reproducibility for the sake of convenience [@Wijesooriya2022-ly].
Transferring data between compute platforms is also discouraged.
For example, having workflows that involve combinations of web-based and scripted tools require
data transfer steps which are inefficient and error-prone.
On the other hand, some web based tools excel at reproducibility.
The web-based analysis platforms Galaxy and GenePattern enable sophisticated point-and-click
bioinformatics analysis in the browser, and those workflows can also be shared in a reproducible way
[@The_Galaxy_Community2022-dt;@Reich2006-na].
Some webtools facilitate reproducibility by providing code (e.g.: Degust @Powell2019-dw)
or by allowing apps to be executed locally (e.g.: ShinyGO @Ge2020-qe).

To facilitate an end-to-end automated process, code and data need to be "linked", which means the
code is aware of the location of the data and how to get it automatically @Peng2011-ia.
The code should be able to fetch the data from a publicly accessible location and conduct the downstream
processing and analysis without having to acquire the data some other way.

@@

A guiding principle of the five pillars approach is that the publishing researchers should strive
to make the reproduction process easier and faster for those who wish to.
A "master script" that coordinates the execution of individual parts of an analysis is an excellent
way to reduce the number of commands required for reproduction.
This can be coded in standard open source languages like R/R Markdown, Python/Jupyter notebook or the
Unix Shell.
These approaches work equally well on personal computers, cloud servers and high performance clusters.
Using free and open source software ensures that the research is accessible to the greatest audience
@Garijo2013-zl, as opposed to proprietary software like SPSS, STATA and MatLab which are cost inhibitory.
Free and open source software also enjoy relatively larger libraries of add-on packages contributed by
the scientific community.
Projects involving computationally intensive tasks might benefit from a build/workflow automation
solution like `make`, `snakemake` [@Koster2012-sq; @Molder2021-mp], `targets` [@Landau2021] or
`nextflow` [@Di_Tommaso2017-yo].
The advantage of such tools is that if the analysis were to terminate due to an issue mid-way,
for example, a hardware problem at step 8 of a 15-step workflow, the analysis of steps 1-7 wouldn't
need to be repeated.
After fixing the hardware issue, re-running the analysis would pick up at step 8 again,
which saves labor and compute time.

@@

Around the same time, Project Jupyter was developed to provide a "notebook" type interface, incorporating
R, Python, Julia and other computer language chunks into documents [@kluyver2016; @granger2021; @Perkel2018-rz].
A noteworthy attempt to use literate programming in Jupyter notebooks to aim for reproducibility
has been the analysis of RNA-seq samples from patients infected by the Zika virus [@wang2016].
In this work, authors provided alongside the traditional research paper a
[Jupyter notebook](https://nbviewer.org/github/maayanlab/Zika-RNAseq-Pipeline/blob/master/Zika.ipynb)
that performed and documented the entire analysis procedure.
The MyST-NB (short for Markedly Structured Text notebook) is built on top of Jupyter notebooks or Markdown documents
and has some notable advantages around flexibility of content that make it an attractive choice for executable
article authoring [@DuPre2022-bk].

@@

* Flexible output formats.
These include PDF, DOC/DOCX and HTML, the latter having some notable benefits
including better usability for mobile devices like phones and tablets, and the ability to
support richer content such as interactive charts and dynamic tables (searchable, filterable, sortable).
Outputs include HTML documents, but also slideshows and even ebooks.
Quarto and R Markdown have the ability to automatically generate a bibliography in many different
journal styles [@noauthor_undated-pg].
Further, thanks to the `rticles` package, dozens of journal style R Markdown document templates are
freely available @Xie2018-rf.

These features make literate programming a useful tool for science communication in a range of
situations.
Whether this is sharing a data analysis report, giving a presentation at a meeting,
writing a research article or self-publishing an ebook, literate programming provides the
ability to construct transparent data narratives with clear provenance in a conveniently shareable
form.

@@

As outlined above, one of the goals is to reduce the complexity of reproduction, in particular by
minimising the number of commands required.
This also makes the `README` documentation much simpler.
For example, the pathway analysis workflow we previously developed could be reproduced and inspected
with just six commands, starting from a new Ubuntu Linux installation [@Ziemann2023-nm].
By minimizing the complexity of reproduction and writing the instructions clearly in the `README`,
reproducibility is made possible to a larger segment of the computational research community,
especially those who are not experts in the languages used.
